---
title: "Simple Statistical Tests -- Class Script"
output: html_notebook
---

```{r 6_load_libraries, message = FALSE}
library(tidyverse)
```

# Lesson Code

First, we'll load and clean up the correlational data set from the Open Stats Lab Google Drive folder.

Note that the `sprintf()` function just replaces text of %s with whatever variable you also provide to the function. So the %s in "id=" becomes the id variable string.

```{r 6_clean_correlation_data}
# variables
scale_max <- 6
id <- "0Bz-rhZ21ShvOMGxnYUJfYmR5d2M" # google file ID

# loading and filtering
corr_data <- read_csv(
  sprintf(
    "https://docs.google.com/uc?id=%s&export=download", 
    id)
  ) %>%
  mutate(redist2 = (scale_max + 1) - redist2,
         redist4 = (scale_max + 1) - redist4) %>%
  transmute(PS, 
            Fair_Satisfied = (fairness + satisfaction) / 2,
            Support = (redist1 + redist2 + redist3 + redist4) / 4
            ) %>%
  rename_all(tolower)
corr_data
```

Make data for paired tests.

```{r 6_generate_paired_data}
set.seed(1000)
stroop_dat <- tibble(
  subject = seq(1:60),
  congruent = rnorm(mean = 400, sd = 30, n = 60),
  incongruent = congruent + rnorm(mean = 30, sd = 10, n = 60)
  ) %>%
  gather(key = condition, value = reaction_time, 2:3) %>% 
  mutate(
    log_RT = log(reaction_time)
  )
stroop_dat
```

Make data for binomial independent samples tests.

```{r 6_binomial_data_generation}
set.seed(1000)
prob <- rep(c(0.3, 0.6), 50)
cond <- rep(c("control", "intervention"), 50)
subject <- seq(1:100)

smoking_dat <- tibble(
  subject = subject,
  cond = cond,
  outcome = rbinom(n = 100, size = 1, prob = prob)
)
```

Add your code chunks using insert, R on the top right panel of this window.

```{r your_code_chunk}

```

