# Data Visualisation 2

In this section we'll cover more advanced plotting methods in ggplot2. We'll look at customising plots, making **pirate plots**, installing packages from GitHub, faceting, and stitching plots together.

To get started, as always we'll load our packages and saved data from the previous lesson.

```{r 3_load_packages, include = FALSE}
library(tidyverse)
```

```{r 3_load_packages_class, eval = FALSE}
library(tidyverse)
```

```{r 3_load_simulated_data, echo = FALSE, message = FALSE}
rt_data <- read_csv("lesson_materials/03_data_visualisation_two/inputs/rt_data.csv")
```

```{r 3_class_load_simulated_data, eval = FALSE}
rt_data <- read_csv("inputs/rt_data.csv")
```

## Customising Your Plots

Take the desnity plot below, this is functional, but it's pretty ugly.

```{r 3_density_plot_boring}
ggplot(data = rt_data, 
       mapping = aes(x = reaction_time, 
                     fill = response_condition
                     )
       ) +
  geom_density(alpha = 0.5)
```

You already know how to change the colours of your aesthetics across all conditions and split by condition, so now we'll look at other ways to improve the plot. 

In this version of the plot, we use the same code as before, but we add labels to the axes using `labs()`, assigning a nicer looking version of our variables to the x and y axes. 

Additionally, we change the scale of the x-axis using the `scale_x_continuous` function. To this function, we pass the **limits** that we want for our axis (between 200 and 600ms), and we identify the breaks, or where we want the ticks along the axis. We pass another function that we learned in the Lesson One, called `seq()`. This sets up a sequence of numbers for us without us having to type them all out. Here, it goes `from` 200 `to` 600 `by` ticks every 100ms; as a result, our axis gets labels of 200, 300, 400, 500, and 600. 

On top of this, we also improved the label for our legend using the `guides()` function. As our legend is only there to identify different parts of the graph with a different colour (from `fill = response_condition` in `aes()`), then we have to tell the guide to change the legend that pops up because of the differnt coloured parts of the plot. So, we change our guide, and change the legend that comes up because of the change in colour (`guides(fill = guide_legend())`) and within the guide legend, we change the title (`title = "Response Condition"`. I know this sounds like a lot to take in, and you're very likely to forget how this works (I do all the time), but hopefully you can get a grip of it by seeing it in action.

Finally, we've changed the theme of our plot to `theme_bw()`. This is one of many inbuilt themes in ggplot2, but I find it one of the cleanest.

```{r 3_density_plot_improved}
ggplot(data = rt_data, 
       mapping = aes(x = reaction_time, 
                     fill = response_condition
                     )
       ) +
  geom_density(alpha = 0.5) +
  labs(x = "Reaction Time (ms)", 
       y = "Density"
       ) +
  scale_x_continuous(limits = c(200, 600), 
                     breaks = seq(from = 200, 
                                  to = 600, 
                                  by = 100
                                  )
                     ) +
  guides(fill = guide_legend(title = "Response Condition")) +
  theme_bw()
```

With the histrogram below, we don't have much new to introduce, expect this time we use `theme_classic()` instead of `theme_bw()`. This gets rid of the major and minor grid lines from the previous plot, and also keeps only the lines for the axes. However, we added the argument `expand` to the `scale_y_continuous()` function, and pass this the values 0 and 0. This makes removes the spacing between the plotted elements and the axes. These values simply indicate how much extra space should be added to the top and bottom of the plot.

```{r 3_histogram_plot}
ggplot(data = rt_data, 
       mapping = aes(x = reaction_time)) +
  geom_histogram(binwidth = 50, 
                 fill = "#bcbddc",
                 colour = "#756bb1"
                 ) +
  scale_x_continuous(limits = c(200, 600),
                     breaks = seq(from = 200, 
                                  to = 600, 
                                  by = 25
                                  )
                     ) +
  scale_y_continuous(expand = c(0, 0)) +
  labs(x = "Reaction Time (ms)", y = "Count") +
  theme_classic()
```

## Pirate Plots

Now we'll look at a new plot type that takes a bit of extra work to generate; pirate plots. Pirate plots are a great choice of plot for the indecisive. They are essentially the individual points of data, a bar plot, a violin plot, and a confidence interval interval all in one. This way, you get to see the raw, descriptive, and inferential data on one plot!

This is a nice way to show data that are grouped by categories but with a continuous dependent variable. 

We could make these ourselves by doing some calculations and combining `geoms` in ggplot2. Or, we could just install a package from [Github](https://github.com/mikabr/ggpirate) which will do that for us. First, we'll have to install `devtools` in R which will allow us to install packages from Github. (For packages on CRAN, this isn't necessary, but unfortunately ggpirate isn't on CRAN at the time of writing.)

To install `devtools` and the `ggpirate` package, uncomment and run the code below. Then as always, use `library(ggpirate)` to load the package.

```{r 3_load_pirate_plot, include = FALSE}
library(ggpirate)
```

```{r 3_install_pirate_plot_class, eval = FALSE}
# install.packages("devtools")
# devtools::install_github("mikabr/ggpirate")
library(ggpirate)
```

Below, we'll make a pirate plot. Note that you only need the first two calls (lines 1-6) to create the plot: The first to set up how you're going to map your data (and the source of your data), and the second to add the `geom` for the pirate plot itself. We added the aethetics of colour and fill to match our conditions within this call, so the two levels of `response_condition` have different colours.

For the additional lines:

- `labs` allows you to manually change the x and y axis labels
- `scale_x_discrete` allows you to manipulate your x scale. Within this, we change the labels of the columns using `labels`. We changed them to Impaired (vs. impaired) and Intact (vs. intact) for stylisitc reasons.
- `scale_y_continuous` allows you to manipulate your y scale. Here, we set the `limits`, i.e. how far to extend our y-axis to between 0 and 600ms. Additionally, we set our `breaks` to increment in the sequence (`seq`) from 0 to 600 by 100ms. This way, we've added more granularity to our axis ticks.
- We use `theme_bw()` to change to a black and white theme. There are other themes in ggplot2, but this one is nice and clean.
- Next, `theme` allows you to manually specify other aspects of how your plot should look. Here, we used `panel.grid.major.x` and set this to nothing (`element_blank()`) because we don't need vertical lines in our plot.
- Finally, we define the colour and fill for our plot using manual hex values using the `scale_colour_manual()` and `scale_fill_manual()` functions.

```{r 3_pirate_plot}
ggplot(data = rt_data, 
       mapping = aes(x = response_condition, 
                     y = reaction_time)
       ) +
  geom_pirate(mapping = aes(colour = response_condition, 
                            fill = response_condition)
              ) +
  labs(x = "Motor Skill", y = "Reaction Time (ms)") +
  scale_x_discrete(labels = c("Impaired", "Intact")) +
  scale_y_continuous(limits = c(0, 600), 
                     breaks = seq(from = 0, to = 600, by = 100)
                     ) +
  theme_bw() +
  theme(panel.grid.major.x = element_blank()) +
  scale_colour_manual(values = c("#af8dc3", "#7fbf7b")) +
  scale_fill_manual(values = c("#af8dc3", "#7fbf7b"))
```

In this plot, we have coloured bars and lines indicating the mean scores, a box representing the 95% confidence interval assuming a normal sampling distribution, violins indicating density, and the raw data points. 

If the 95% confidence interval between the two groups doesn't overlap, we can be fairly sure there is a significant difference between the groups. So, here we can be fairly certain the two groups differ in reaction times.

## Faceting

Another useful part of plotting in ggplot2 is that you can make facets of plots, or subplots. This is a good way to display your data if you have multiple categorical variables. Essentially, you'll get a plot for each category in your data.

### Facet Wrap

If you want to create facets from one variable then use `facet_wrap()`.

```{r 3_facet_wrap}
ggplot(data = rt_data, mapping = aes(x = reaction_time)) +
  geom_histogram(binwidth = 50, 
                 fill = "white", 
                 colour = "black"
                 ) +
  facet_wrap(~ response_condition)
```

In this plot, we've specified a histogram as we normally would. However, we use `facet_wrap()` and a tilde (`~`) to create a formula for how to display our plots. We define our variable with which to split our plots to the right of the `~`, and ggplot automatically plots the two separately at the same time. Notice that we get useful labels at the top of these plots, too.

### Facet Grid

If we wanted to make a facet by two variables, then we would use `facet_grid()` instead. In this case, we just add each variable to either side of the `~` and ggplot will do the splitting for us. 

Let's see how this works if we split our data by gender and response condition.

```{r 3_facet_grid}
ggplot(data = rt_data, mapping = aes(x = reaction_time)) +
  geom_histogram(binwidth = 50, 
                 fill = "white", 
                 colour = "black"
                 ) +
  facet_grid(gender ~ response_condition)
```

The order in which you specify the two variables matters. Try swapping around between `facet_wrap(gender ~ response_condition)` and `facet_wrap(response_condition ~ gender)` to see how this works.

## Calculating Statisitcs in ggplot2

Sometimes, plotting just the means isn't enough. We saw how useful the 95% confidence interval from `ggpirate` is for making inferences about the differences between groups. Nicely, we can get standard errors or confidence intervals around our data points within ggplot for other geoms. 

### Means and Error Bars

Let's say you wanted a bar plot with error bars showing the standard error. You can create this in ggplot using the `stat_summary()` function. In the first instance here, we tell it that we want to run the function `mean` over our data that make up the y-axis; hence `fun.y = mean`. We also need to specify which geom we want to return from this. 

Try changing the geom to `point` in the first `stat_summary()` call to see what happens when you run this plot with `geom = "point"`. 

Finally, we ask for another summary, but this time we want an error bar. So, for the geom call we request an error bar. Crucially, the function we require to get this errorbar is `fun.data = "mean_se"`. That's because we need the mean to know where the centre the errorbar, and the standard error to get the limits of the bar. We manually changed the width of the bar to a quarter of the bar size using the `width` argument to stop ggplot returning super wide error bars.

```{r 3_error_bars}
ggplot(data = rt_data, 
       mapping = aes(
         x = response_condition,
         y = reaction_time, 
         fill = response_condition
         )
       ) + 
  stat_summary(fun.y = "mean", geom = "bar") +
  stat_summary(fun.data = "mean_se", geom = "errorbar", width = 0.25)
```

I don't often use `stat_summary` in my own plots, as I often want to know exactly how I've calculated things like the standard error. Doing things by hand allows you to change error bar sizes appropriately for within- and between-subject designs. But, this is useful if you want to create a plot rapidly, or want to avoid the hassle of writing extra code.

### Model Fits

Here' we'll switch again to a different data set that has two continuous variables. The `starwars` data set is useful for this exercise.

We can use the `geom_smooth()` function to fit a model to our data. This defaults to a loess method, but we can change this to a linear model (or other alternatives) as in the second plot below. By default, these smoothed plots display a ribbon around the fit which indicates the confidence interval (95% by default).

```{r 3_smooth_starwars}
# remove NA and mass above 1000
filtered_starwars <- starwars %>% 
  drop_na(height, mass) %>% 
  filter(mass < 1000)

# plot
ggplot(data = filtered_starwars, 
       mapping = aes(x = mass, y = height)
       ) + 
  geom_point() + 
  geom_smooth()
```

Next, we'll change the defaults in order to fit a linear model. We do this in the `geom_smooth` function, `method = "lm"`. We can additionally specify a function, so we can change how we fit the data. Currently, our formula is `y ~ x`, which is a regular linear model. We could, however, fit a quadratic function to the data by using `y ~ poly(x, 2)`. The function `poly(x, 2)` calculates orthogonal polynomials to a certain degree. The first being linear, second quadratic (think curved lines with 1 inflection point), the third cubic (think curved lines with 2 inflection points), etc.. For now, we'll just fit the linear model.

```{r 3_linear_starwars}
ggplot(data = filtered_starwars, 
       mapping = aes(x = mass, y = height)
       ) + 
  geom_point() + 
  geom_smooth(method = 'lm', formula = y ~ x)
```

Alternatively, if we have a fitted model saved from our analyses, we can add the fitted line directly using the `stat_summary` function. However, this requires some insight into the models fitted, so we'll save this for later classes.

Finally, we'll look at how we can combine these smoothed fits with averages of scores. This is often useful if you're looking at timecourse data and you want to summarise performance across a whole range of participants. We'll also look at adding a few of the graphical flourishes from above to make the plot look really nice.

Next, we'll look at making the point and line points look a little better. For this, we'll use the inbuilt `ChickWeight` data set from R. This is similar to the `chickwts` data set from the previous session, but crucially this data is not aggregaged across time. Instead, we have several measurements for the chicks at different time points. This makes this data especially nice for plotting changes over time in our sample.

First, load the data.

```{r 02_load_ChickWeight_data}
data(ChickWeight)
```

Then we can see how the data looks. This data is stored as a data.frame, so we have to be careful when printing it out. We'll use the `head()` function to get only the first 6 rows of data.

```{r 02_head_ChickWeight_data}
head(ChickWeight)
```

We have 4 columns: 

- weight: out dependent variable, the weight of the chicks
- Time: an independent variable indicating the time point at which a measurement was taken
- Chick: an ID column for each chick
- Diet: an independent variable indicating the diet of the chick

Unfortuantely, we only know that there are different diets from the data set, but not what those diets are. We also don't know the units by which weight is measured or by which the measurements were taken. 

This is a good lesson in using sensible variable names. If you want yourself and others to understand your data years down the line with little background knowledge of your study, then use informative labels for your data!

We'd expect that as time goes on, chick weights should increase. Also, we might expect that the rate of growth should differ for each diet. As such, we'll plot this together using our data set. 

First, we'll plot our chick data, using the `stat_summary()` function to plot some mean points for the chick weights, and the `geom_smooth()` function to fit a linear model to the data. We'll make sure that the points, lines, and ribbons (indicating the confidence interval) are different colours for each level of the diet factor.

We're going to build up our plot bit by bit, so this time we can same our plot as an object, before returning the object to see the plot.

```{r 3_continuous_data_simple}
chick_time_plot <- ggplot(data = ChickWeight, # data
                          mapping = aes(x = Time, # x-axis
                                        y = weight, # y-axis
                                        colour = Diet, # colour ID
                                        fill = Diet # fill ID
                                        )
                          ) +
  stat_summary(fun.y = mean, geom = "point") + # point means
  geom_smooth(method = 'lm', formula = y ~ x) # linear fit

# return the plot
chick_time_plot
```

We can improve this plot. So, let's add some further information. What if we had a target weight we'd like our chickens to reach over time, and we want to see when each diet pushes chicks over this line? 

We can add this information by including a horizontal line using `geom_hline()`. We just need to specify where on the y-axis this line should go (`yintercept`), and which style we want for the line (`linetype`). We should also add an annotation to the line indicating what it represents. We can do this by using the `annotate` function, and specifying that we want a text geom. We need to specify the x and y values for where the text lies, and what the text should say.

To add this to our original plot, we'll overwrite the original plot name with itself, plus our additional arguments.

```{r 3_continuous_data_pretty}
chick_time_plot <- chick_time_plot +
  geom_hline(yintercept = 150, linetype = 2) + # horizontal line
  annotate(geom = "text", x = 2, y = 158, label = "Target Weight") # line label

# return the plot
chick_time_plot
```

Finally, we can improve the general look of the plot by adding some additional arguments. We'll specify the labels for the x and y axes using the `labs()` function. 

We'll also specify the limits of the y-axis using the `coord_cartesian()` function and specifying the minimum and maximum values for the limits of the y-axis.

Next, we'll specify a general basic theme for the plot using the `theme_bw()` argument. We'll change a minor detail of this theme, namely the position and style of the legend by specifying additional `theme()` arguments. This has to come after `theme_bw()` so `theme_bw()` doesn't overwrite our additional arguments. First, we simply state the x and y locations for the legend by defining the `legend.position`. Next, we want a black border around our legend, so we do this by changing the `legend.background`. Here, we set colour (the outline) to black, with a relatively small line size (1), and a solid linetype.

Finally, we can change the title of our legend if we'd like. The original title is quite informative, but it's useful to know how to do this. Here, we use the `guides` function, and specify both the guide legend title using `guide_legend()` for both the `colour` and `fill` properties. It's important to do this for both properties as our data is identified by colours on both of these properties. If we just change the title of one of the properties (e.g. fill), then we'll have two legends! 

```{r 3_continuous_data_prettiest}
chick_time_plot <- chick_time_plot +
  labs(x = "Trial", y = "Weight") + # axis labels
  coord_cartesian(ylim = c(0, 300)) + # range of y-axis
  theme_bw() + # general theme
  theme(legend.position = c(x = 0.15, y = 0.8), # position of legend
        legend.background = element_rect(color = "black",
                                         size = 1,
                                         linetype = "solid"
                                         ) # legend styling
  ) +
  guides(colour = guide_legend("Experimental Diet"),
         fill = guide_legend("Experimental Diet")
         ) # legend title

# return the plot
chick_time_plot
```

## Combining Plots

Finally, while it's all well and good plotting one model at a time, often if you're writing a paper you want several plots together to save space. You could do this by hand in certain image editing packages, but luckily there's a package for this on CRAN; `cowplot`. First, we need to install and load the package. Intall this by running the commented out code below. After that, load `cowplot` each time you want to use it. 

```{r 3_load_cowplot, message = FALSE}
# install.packages("cowplot")
library(cowplot) # run this each time
```

In order to stitch two plots together, we need to save our plots as objects so they're available to the R environment. This is the same process we used above for the `ChickWeight` plots.

If we assign our plotting code to an object, then every time we run the name of the object, we'll get the output of that code back.

For keeping several plots in my environemnt at once, I often save them to an object. In this instance, we'll save a plot of just the points to the name `point_plot`. That's because we want to overlay a linear fit and a quadratic fit on this plot separately, but we don't want to type the full code out twice. Instead, we fit the plot of points to `point_plot`, then add a linear or quadratic fit by adding the `geom_smooth` argument to `point_plot` and saving both under new names. We can do this like so:

```{r 3_save_linear_plot}
# create a plot of points
point_plot <- ggplot(data = filtered_starwars, 
                     mapping = aes(x = mass, y = height)
                     ) +
  geom_point() +
  coord_cartesian(ylim = c(0, 300))

# create a linear plot by adding the fit to the plot of points
linear_plot <-  point_plot + 
  geom_smooth(method = 'lm', formula = y ~ x)
```

When you do this, you won't automatically see the plot once you run the code unless you run the object assigned to the plot. Let's try this for a quadratic fit of the same data.

```{r 3_save_quadratic_plot}
# create a quadratic plot by adding the fit to the plot of points
quadratic_plot <- point_plot + 
  geom_smooth(method = 'lm', formula = y ~ poly(x, 2)) # fit quadratic
  
# return the plot
quadratic_plot
```

You can see that we've got the quadratic fit and the 95% confidence interval around this fit from the above code. Why does the plot look different to the base plots in ggplot2? `cowplot` loads some defaults for all plots outputted by `ggplot` to save you on typing out your own theme arguments.

Now, we can combine these two plots into a single plot using the new functionalities from `cowplot`.

```{r 3_combine_plots}
combined_plots <- plot_grid(linear_plot, 
                            quadratic_plot, 
                            labels = c("A", "B"), # label plots
                            align = "h" # align axes
                            ) 
combined_plots
```

What if we want a title for our plots? We first have to define our title with a combination of `ggdraw()` and `draw_label()`. Inside `draw_label()` we used a new function, `paste()` to paste two strings of text together (with a space between the two strings). We could simply input the entire string, but we broke it down into two bits so we don't exceed the 80 character width limit from our style guide! 

We further specify a bold `fontface` in the `draw_label()` command outside our pasted title.

Finally, we display the plot by creating a `plot_grid()` of our plot and the title in this order so the title displays underneath the plot. We specify that we just want 1 column so the plot and title are stacked together, and we specify the relative heights of the title and plot separately so that the title is smaller than the plot so it doesn't take up an equal amount of space as the plot.

```{r 3_add_plot_title}
# create title
title <- ggdraw() + 
  draw_label(paste("Linear (A) and Quadratic",
                   "(B) fits of Height and Weight"
                   ), 
             fontface = "bold"
             )

# print plot as a grid
combined_plots <- plot_grid(combined_plots, 
                            title, 
                            ncol = 1, 
                            rel_heights = c(1, 0.1)
                            )

# return the combined plots
combined_plots
```

You can find further information on adding joint titles, annotations, etc. in [this article](https://htmlpreview.github.io/?https://github.com/wilkelab/cowplot/blob/master/inst/doc/plot_annotations.html) by Claus O. Wilke.

Note that you can reset the the default styles inherited by cowplot so that your plots return to their defaults with the following code:

```{r 3_cowplot_reset_defaults}
theme_set(theme_gray())
```


## Saving Plots

Finally, if you've went to the work of producing a nice plot, then it'll be useful to know how to save it somewhere. To save our combined plot of the linear and quadratic fits, we'll use `ggsave()`. You can name your plot anything you like, but remember to add a file extension at the end. I've used .png as this format suffers from fewer artifacts when comapred to JPEG, and it's a pretty common filetype.

```{r 3_save_plots_class, eval = FALSE}
ggsave(filename = "outputs/starwars_mass_by_height.png", 
       plot = combined_plots
       )
```

You can do a whole bunch of other things with ggplot, like adding vertical and horizontal lines (often useful for identifying chance performance in participants), and annotating plots directly (useful for adding statistics, or commenting on interesting sections of the data). We'll cover these in the exercises, however, as these are just additional flavour to our plots!

## Exercises

### Main Exercises

If all this complexity hasn't given you [nausea](https://youtu.be/JidJV1ue1lI), try out the exercises below. We'll cover these in the class with the solutions uploaded at the beginning of each session in a separate downloadable link. Try to solve these questions before resorting to the solutions. I'll be there to help out where necessary.

### Question 1

Let's assume we want to know if the density of the mean reaction times differs across word class within our items.

Using the `lex_dec_items` data set, make a density plot of the word frequency by class. Save this plot as `lexical_density`, and output the plot.

### Question 2

Add a black and white theme to your `lexical_density` plot. Next, give the axis labels and legend labels uppercase names. Finally, give your legend title the name Word Class. Assign all of this to a new object `lexical_density_improved`, and output your plot below.

*Note*: To change the legend labels, you need to use both `scale_fill_discrete` and `scale_colour_discrete`. Why do you think this is? Why don't we just use `scale_x_discrete` as we did in class?

### Question 3

There's some repetition in the code for the plot above. Can you improve your code to remove that?

### Question 4

Now we want to check the distribution of the reaction times depending on word class and the language spoken by our participants.

Using the `lex_dec` data set, create a faceted plot that looks at the density of reaction times. This should be made up of a grid of densities split by native language and word class.

Assign this to the object `rt_density` and output your plot.

### Question 5

Now we want to explore if there's any relationship between the mean reaction time to our items and the frequency of the item.

Using the `lex_dec_items` data set, plot the relationship between word frequency and mean reaction time as a scatter plot. We want a fitted line and points for the mean reaction time.

### Question 6

Now we want to know how many males and females took part in our experiment.

Using the `lex_demog` data set, create a count of the number of males and females who took part in the experiment. Make all text in the plot uppercase, and make the plot theme black and white.

Assign this to the object `gender_count` and output your plot.

### Question 7

What if we want to know the mean age and distribution of ages split by language spoken and gender?

Using the `lex_demog` data set, create a pirate plot of the ages by each gender. You can set the colour and fill to gender to have more colourful plot if you'd prefer.

Additionally, we would like these plots split by the native language of the speaker, so facet your plot by native language.

Assign this all to the object `demographic_age`, and output your plot.

Can you see how a bar plot of the average ages might be misleading in this instance? Pay particular attention to the male bar for the other language group.
 
### Question 8

Next, we want to see the mean (and standard error) of reaction times to words with different word frequencies.

Using the `lex_dec` data set, create a scatter plot of the reaction time by word frequency. We would like this split by word class. 

*Hint*: Be sure to use the `stat_summary` function to get pointranges that represent the mean and standard error.

Assign this to the object `rt_pointrange`, and output your plot.

### Question 9

Finally, we want to show a few of our graphs in one display.

Using the cowplot library, stitch together the plots for question 6, 7, 4, and 8. 

Add the labels A-D to identify each plot.

Save these plots under the object `combined_plots` and return your combined plots.

### Additional Exercise

We can improve the combined plots above.

Add a (short) simple title to your combined plots, and save this plot in the outputs folder.

These plots won't look perfect, and you may need to change the font and element sizes. That can be easily achieved, but we won't do that here.
