---
title: "05 - Data Manipulation 2"
subtitle: data transformation and summaries
author: Glenn Williams
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: ioslides_presentation
---

<style>
pre {
  width: 100%; /* 106%; */
  left: 0; /* -60px; */
  padding: 10px 15px 10px 15px; /* 10px 0 10px 60px; */
  font-size: 18px;
}
</style>

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(fig.width = 5, fig.height = 3) 
```

# Overview

## Lesson Structure

This lesson We'll look at these functions, that work with `group_by()`:

- `arrange()`: order variables
- `select()`: pick out variables
- `filter()`: pick out observations
- `mutate()`: create new variables
- `summarise()`: create a summary of variables (e.g. mean, *SD*, n)

We'll also cover `rename()` for renaming variables.

## Getting Started

First, we'll load the packages necessary for this class. If you don't have lubridate, install it with `install.packages("lubridate")`.

```{r 3_load_packages, message = FALSE}
library(tidyverse)
library(lubridate)
```

Then load your data (already done in your class file).

```{r 5_load_demo_data, include = FALSE}
demo <- read_csv("../inputs/lexical_decision_demographic_data_wide.csv")
```

```{r 5_load_demo_data_class, eval = FALSE}
demo <- read_csv("inputs/lexical_decision_demographic_data_wide.csv")
```

## Understanding our Data

`glimpse()` trasposes data (with data type): see the first few observations of every column

```{r 5_glimpse_demo_data}
glimpse(demo)
```

## Preparing our Data

Let's separate **funRec** and **completion_time** using the `separate()` function from **tidyr**. This sort of tidying is common (see previous lesson).

```{r 5_glimpse_demo_data_separated_code, eval = FALSE}
demo <- demo %>%
  separate(col = funRec, into = c("fun", "recommend")) %>%
  separate(col = completion_time, into = c("start", "end"), sep = "_")
glimpse(demo)
```

## Preparing our Data

```{r 5_glimpse_demo_data_separated, echo = FALSE}
demo <- demo %>%
  separate(col = funRec, into = c("fun", "recommend")) %>%
  separate(col = completion_time, into = c("start", "end"), sep = "_")
glimpse(demo)
```

## Selecting Columns

Can select by name or number.

```{r 5_drop_programming_columns_name_code, eval = FALSE}
# by name
demo %>% select(ID, 
                LANGUAGE, 
                progress, 
                gender, 
                age, 
                tester, 
                fun,
                recommend,
                start,
                end
                )

# by number
demo %>% select(c(1, 8:16))
```

## Selecting Columns

Can select by name or number.

```{r 5_drop_programming_columns_name, echo = FALSE}
# by name
demo %>% select(ID, 
                LANGUAGE, 
                progress, 
                gender, 
                age, 
                tester, 
                fun,
                recommend,
                start,
                end
                )

# by number
demo %>% select(c(1, 8:16))
```

## Selecting Columns

Can also specify columns to drop (numbers 2 through to 7) with a minus

```{r 5_drop_programming_columns_number}
demo_sub <- demo %>% select(-c(2:7))
demo_sub
```

## Selecting Columns

We have many select helpers, but we won't cover this here. Try to mess around with these helpers in your own time:

- `starts_with("string")`: keeps columns starting with "string"
- `ends_with("string")`: keeps columns ending with "string"
- `contains("string")`: keeps columns with names containing "string"
- `matches(regular_expression)`: keeps columns matching a **regular expression**
- `num_range("prefix", range)`: keeps columns with a matching prefix and a following range of numbers

## Renaming Columns

Takes the form, `rename(new name = old name)`

```{r 5_rename_columns}
demo_sub <- demo_sub %>% rename(language = LANGUAGE)
demo_sub
```

## Reordering Columns

Use `select()` to reorder columns by specifying columns to keep in order. `everything()` helps if you want to keep the same original order after things you've specified.

```{r 5_select_columns_and_everything}
demo_sub %>% select(ID, age, gender, everything())
```

## Creating and Changing Columns

- `mutate()` allows you to create a new column
- You can do this from manipulations to old columns (or not)
- Set a column to a mutated version of itself to change a column, e.g.:
- Can refer to just-created columns to make new columns 

```{r 5_mutate_columns_completion_time_code, eval = FALSE}
# requires lubridate for ymd_hms() function
demo_sub <- demo_sub %>%
  mutate(start = ymd_hms(start),
         end = ymd_hms(end),
         time = end - start
         )
glimpse(demo_sub)
```

## Creating and Changing Columns

```{r 5_mutate_columns_completion_time, echo = FALSE}
# requires lubridate for ymd_hms() function
demo_sub <- demo_sub %>%
  mutate(start = ymd_hms(start),
         end = ymd_hms(end),
         time = end - start
         )
glimpse(demo_sub)
```

## Creating and Only Keeping New Columns

`transmute()` is like `mutate()` but only keeps the new cols.

```{r 5_transmute_columns_code_class, eval = FALSE}
# calculate difference from mean completion time
transmuted_time <- demo_sub %>%
  mutate(
    start = ymd_hms(start),
    end = ymd_hms(end),
    time = end - start
  ) %>%
  transmute(
    ID,
    time_diff = time - mean(time)
  )

units(transmuted_time$time_diff) <- "mins" # define units
transmuted_time # print the object
```

## Creating and Only Keeping New Columns

`transmute()` is like `mutate()` but only keeps the new cols.

```{r 5_transmute_columns_code, include = FALSE}
# calculate difference from mean completion time
transmuted_time <- demo_sub %>%
  mutate(
    start = ymd_hms(start),
    end = ymd_hms(end),
    time = end - start
  ) %>%
  transmute(
    ID,
    time_diff = time - mean(time)
  )

units(transmuted_time$time_diff) <- "mins" # define units
transmuted_time # print the object
```

```{r 5_transmute_columns}
transmuted_time # print the object
```

## Filtering to Observations

How many labels do we have for progress?

```{r 5_unique_labels}
unique(demo_sub$progress)
```

- We have a few labels: FINISH, no, END, ethics, and NA.
- Filter: **FINISH** and **END**

## Filtering with Logical Operations

These two are equivalent. 

1. filters where progress values ares equal to (`==`) "FINISH" *or* (`|`) END

Use the second when you have many OR statements.

```{r 5_filter_not_completion_one}
demo_sub %>% filter(progress == "FINISH" | progress == "END")
```

## Filtering with Logical Operations

These two are equivalent. 

2. filters where FINISH or END values are in (`%in%`) progress column

Use the second when you have many OR statements.

```{r 5_filter_not_completion_two}
demo_sub %>% filter(progress %in% c("FINISH", "END"))
```

## Combining Filtering Criteria

Add filters by other columns with a comma. You can use different operators to filter.

```{r 5_filter_not_completion_many}
demo_sub %>% filter(progress %in% c("FINISH", "END"),
                    tester == "RHB",
                    age > 30
                    )
```

## Removing by Ceriteria

Which columns should we **not** keep? Use `!` (read: not) before criteria. 
- This only gives you people who didn't finish.

```{r 5_not_filtering}
demo_sub %>% filter(!progress %in% c("FINISH", "END"))
```

## Handling NAs

This won't run...

```{r 5_incorrect_NA_filtering, eval = FALSE}
demo_sub %>% filter(language == NA)
```

Instead, use `is.na()`, to check whether these values are NAs.

```{r 5_correct_NA_filtering}
demo_sub %>% filter(is.na(language))
```

## Handling NAs

Default filtering = exclude things that don't meet criteria, and NAs.

Here we will filter those who finished the experiment, or who we're unsure about. 

```{r 5_correct_NA_filtering_with_others}
demo_sub %>% filter(progress == "FINISH" | progress == "END" | is.na(progress))
```

## Handling NAs

Or we could do the opposite, dropping these observations.

```{r 5_correct_NA_filtering_with_others_opposite}
demo_sub %>% filter(!progress %in% c("FINISH", "END") & !is.na(progress))
```

## Arranging Data

`arrange()` by any number of variables. Goes from smallest to largest values.

```{r 5_arrange_simple}
demo_sub %>% arrange(ID)
```

## Arranging Data

Arrange by tester, then by start time within tester.

```{r 5_arrange_start_tester}
demo_sub %>% arrange(tester, start)
```

## Arranging Data

Largest to smallest, using the `desc()` function (read: descending):

```{r 5_arrange_desc}
demo_sub %>% arrange(desc(ID))
```


## Summarising Data

- `summarise()` collapses across all observations and creates a table.
- You only get columns you request; here we make the column mean_time by calculating `mean()` on time

```{r 5_summarise_time}
demo_sub %>% summarise(mean_time = mean(time))
```

## Summarising Data

We can add more columns by separating them with a comma:

```{r 5_summarise_time_sd_N}
demo_sub %>% 
  summarise(mean_time = mean(time),
           sd_time = sd(time),
           N = n()
           )
```

## Summarising Data

`summarise()` doesn't play well with NAs.

```{r 5_summarise_time_sd_N_with_NA}
demo_sub %>% 
  mutate(time = replace(time, 1, NA)) %>%
  summarise(mean_time = mean(time),
           sd_time = sd(time),
           N = n()
           )
```

## Summarising Data

`summarise()` doesn't play well with NAs. So we have to tell it to ignore them.

```{r 5_summarise_time_sd_N_with_NA_removed}
demo_sub %>% 
  mutate(time = replace(time, 1, NA)) %>%
  summarise(mean_time = mean(time, na.rm = TRUE),
           sd_time = sd(time, na.rm = TRUE),
           N = n()
           )
```

## Grouping Data

- `group_by` allows you to perform the previous operations on groups of data.
- Most often used with `summarise()` for group summaries

```{r 5_summarise_time_sd_N_no_NA}
demo_sub %>% 
  group_by(language) %>%
  summarise(mean_time = mean(time, na.rm = TRUE),
           sd_time = sd(time, na.rm = TRUE),
           N = n()
           )
```

## Grouping Data

Identifies unique observations in your grouped column, so be sure you pass the right information along!

```{r 5_to_lower_summarise}
demo_sub %>% 
  mutate(language = tolower(language)) %>%
  group_by(language) %>%
  summarise(mean_time = mean(time, na.rm = TRUE),
           sd_time = sd(time, na.rm = TRUE),
           N = n()
           )
```

## Grouping Data

- Works with other functions, like `filter()`
- Here, we remove the lowest time (rank = 1) from each group

```{r 5_to_lower_group_filter}
demo_sub %>% 
  mutate(language = tolower(language)) %>%
  group_by(language) %>%
  filter(rank(time) == 1) %>%
  glimpse()
```

## Ungrouping Data

Use `ungroup()` if you want to go back to ungrouped data.

```{r 5_ungroup_summary}
demo_sub %>% 
  mutate(language = tolower(language)) %>%
  group_by(language) %>%
  filter(rank(time) != 1) %>%
  ungroup() %>%
  summarise(mean_time = mean(time, na.rm = TRUE),
            sd_time = sd(time, na.rm = TRUE),
            N = n()
            )
```

## Chaining Many Functions

We can do all our data cleaning in 1 chain (or by several pipes)

```{r 5_all_together_class, eval = FALSE}
# load, clean, and transform data
demo_clean <- read_csv("inputs/lexical_decision_demographic_data_wide.csv") %>%
  separate(col = funRec, into = c("fun", "recommend")) %>%
  separate(col = completion_time, into = c("start", "end"), sep = "_") %>%
  select(c(1, 8:16)) %>%
  rename(language = LANGUAGE) %>%
  mutate(start = ymd_hms(start),
         end = ymd_hms(end),
         time = end - start,
         language = tolower(language)
         )
```

```{r 5_all_together, include = FALSE}
# load, clean, and transform data
demo_clean <- read_csv("../inputs/lexical_decision_demographic_data_wide.csv") %>%
  separate(col = funRec, into = c("fun", "recommend")) %>%
  separate(col = completion_time, into = c("start", "end"), sep = "_") %>%
  select(c(1, 8:16)) %>%
  rename(language = LANGUAGE) %>%
  mutate(start = ymd_hms(start),
         end = ymd_hms(end),
         time = end - start,
         language = tolower(language)
         )
```

## Chaining Many Functions

Pass this along to any summary code

```{r 5_clean_summary}
demo_clean %>%
  group_by(language) %>%
  summarise(mean_time = mean(time, na.rm = TRUE),
           sd_time = sd(time, na.rm = TRUE),
           N = n()
           )
```

## Chaining Many Functions

Pass this along to any plotting code

```{r 5_clean_plot, message = FALSE}
demo_clean %>%
  ggplot(mapping = aes(x = language, y = time)) + 
  geom_boxplot()
```

## Saving your Data

### CSVs

Save as a .csv for platform-agnostic data

```{r 5_save_demo_data_csv_class, eval = FALSE}
write_csv(demo_clean, "outputs/filtered_demographic_data.csv")
```

### RData Files

Keeps data types, less storage space, allows advanced data layouts (e.g. nested, model outputs)

```{r 5_save_demo_data_Rdata_class, eval = FALSE}
save(demo_clean, file = "outputs/filtered_demographic_data.RData")
```

Load it back up in another session with 

```{r 5_load_demo_data_Rdata_class, eval = FALSE}
load("outputs/filtered_demographic_data.RData")
```

## A Final Note

There are many variants of the functions we've used. Use `?function name` (e.g. `?mutate()`) to find out more about them. Often, you can save typing if you want to mutate several columns using some variant of `mutate()`.

Examples:

- `summarise_if()`, `mutate_if()`, `transmute_if()`: operations on columns where if statement returns TRUE (e.g. `if(is.numeric())`)
- `summarise_at()`, `mutate_at()`, `transmute_at()`: use `select()` helpers to grab column names
- `summarise_all()`, `mutate_all()`, `transmute_all()`: apply functions to all non-grouped columns

# Exercises
