---
title: "04 - Data Manipulation 1"
subtitle: data tidying, joins, and binds
author: Glenn Williams
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: ioslides_presentation
---

<style>
pre {
  width: 100%; /* 106%; */
  left: 0; /* -60px; */
  padding: 10px 15px 10px 15px; /* 10px 0 10px 60px; */
  font-size: 18px;
}
</style>

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(fig.width = 5, fig.height = 3) 
```

# Overview

## Lesson Structure

This lesson We'll look at:

- Loading Data (why `read_csv()`?)
- Converting between long and wide formats: `gather()` vs. `spread()`
- Splitting columns with `separate`
- Joins: mutating, filtering, and binding
- Checking for duplicate entries across multiple tables

## Getting Started

First, we'll load the packages necessary for this class. 

```{r 3_load_packages, message = FALSE}
library(tidyverse)
```

- We'll use the **tidyr** package from the tidyverse to clean and transform our data. 
- We'll also use the **dplyr** package from the tidyverse to join different sets of data together.

## Data Formats

I've saved the data in both **wide** and **long** formats. 

- Wide: each row represents one participant, several observations go in separate columns
- Long: each column represents one measurement, with several observations in separate rows

## Loading Data

```{r 4_load_demo_data, include = FALSE}
demo_wide <- read_csv("../inputs/lexical_decision_demographic_data_wide.csv")
demo_long <- read_csv("../inputs/lexical_decision_demographic_data_long.csv")
```

```{r 4_load_demo_data_class, eval = FALSE}
demo_wide <- read_csv("inputs/lexical_decision_demographic_data_wide.csv")
demo_long <- read_csv("inputs/lexical_decision_demographic_data_long.csv")
```

Use `read_csv()`: parses data as a `tibble` and guesses data types.

## Wide Data

```{r 4_wide_data}
head(demo_wide)
```

- Lots of NA values; if someone doesn't know that language, they can't have a value.
- NA is evidence of absence; avoid blank cells (absence of evidence).

## Long Data

I've cut out a few columns here just so you see the `computer_language` column. But in an R notebook you can just scroll along.

```{r 4_long_data_class, eval = FALSE}
demo_long
```

```{r 4_long_data, echo = FALSE}
demo_long[, c(1: 5, 9)]
```

## Converting from Wide to Long Data with `gather()`

We use the `gather()` function to put multiple columns together in one. This requires:

- Key: name to call the ID column
- Value: name to store values related to the IDs
- IDs for columns to group (by number or name)

## Converting from Wide to Long Data with `gather()`

```{r 4_gather_demo_data_class, eval = FALSE}
gather(data = demo_wide,
       key = prog_lang,
       value = known,
       2:7
       )
```

```{r 4_gather_demo_data, echo = FALSE}
gather(data = demo_wide,
       key = prog_lang,
       value = known,
       2:7
       ) %>%
  select(c(1:6, 9:10))
```

## The Pipe

We can add other functions, like ordering in a readable way with the `magrittr` pipe.

```{r 4_gather_demo_data_pipe_class, eval = FALSE}
demo_wide %>% 
  gather(
    key = prog_lang,
    value = known,
    2:7
    )
```

```{r 4_gather_demo_data_pipe, echo = FALSE}
demo_wide %>% 
  gather(
    key = prog_lang,
    value = known,
    2:7
    ) %>%
  select(c(1:6, 9:10))
```

## The Pipe: Adding Functions

```{r 4_gather_arrange_demo_data_pipe_class, eval = FALSE}
demo_wide %>% 
  gather(
    key = prog_lang,
    value = known,
    2:7
    ) %>%
  arrange(ID)
```

```{r 4_gather_arrange_demo_data_pipe, echo = FALSE}
demo_wide %>% 
  gather(
    key = prog_lang,
    value = known,
    2:7
    ) %>%
  arrange(ID) %>%
  select(c(1:6, 9:10))
```

## Removing NAs from Gathered Data

```{r 4_gather_arrange_select_demo_data_pipe_class}
demo_gathered <- demo_wide %>% 
  gather(
    key = prog_lang,
    value = known,
    2:7,
    na.rm = TRUE
    ) %>%
  arrange(ID) %>%
  select(-known)
```

- `na.rm = TRUE`: removes NA values (read: remove NAs? True (yes))
- `select(-known)`: removes the known column (redundant) (read: pick everything out except known)

## Removing NAs from Gathered Data

```{r 4_gather_arrange_select_demo_data_pipe_class_see_class, eval = FALSE}
demo_gathered
```

```{r 4_gather_arrange_select_demo_data_pipe_class_see, echo = FALSE}
demo_gathered %>% select(c(1:6, 9))
```

## Separating Columns

We have messy data; two variables in one column. How to fix this? `separate()`:
- col = column to split (with bare column name)
- into = into which columns? (needs to be `c("name_one", "name_two")`)

## Separating Columns

```{r 4_separate_funrec_column_class, eval = FALSE}
demo_gathered %>%
  separate(
    col = funRec,
    into = c("fun", "recommend")
    )
```

```{r 4_separate_funrec_column, echo = FALSE}
demo_gathered %>%
  separate(
    col = funRec,
    into = c("fun", "recommend")
    ) %>% 
  select(c(1, 6:9))
```

## Separating Columns: Converting Data Types While we Go!

- `convert = TRUE`!

```{r 4_separate_funrec_column_convert_class, eval = FALSE}
demo_gathered %>%
  separate(
    col = funRec,
    into = c("fun", "recommend"),
    convert = TRUE
    )
```

```{r 4_separate_funrec_column_convert, echo = FALSE}
demo_gathered %>%
  separate(
    col = funRec,
    into = c("fun", "recommend"),
    convert = TRUE
    ) %>% 
  select(c(1, 6:9))
```

## Piping Multiple Arguments

```{r 4_separate_funrec_datetime_column_broken_class, eval = FALSE}
demo_gathered %>%
  separate(
    col = funRec,
    into = c("fun", "recommend"),
    convert = TRUE
    ) %>%
  separate(
    col = completion_time,
    into = c("start_time", "end_time")
    )
```

## Piping Multiple Arguments

```{r 4_separate_funrec_datetime_column_broken, echo = FALSE}
demo_gathered %>%
  separate(
    col = funRec,
    into = c("fun", "recommend"),
    convert = TRUE
    ) %>%
  separate(
    col = completion_time,
    into = c("start_time", "end_time")
    ) %>% 
  select(c(1, 6:10))
```

## Splitting Problematic Columns: Specify Separations

- `sep = "_"`: separate colum values at underscores

```{r 4_separate_funrec_datetime_column_fixed}
demo_gathered <- demo_gathered %>%
  separate(
    col = funRec,
    into = c("fun", "recommend"),
    convert = TRUE
    ) %>%
  separate(
    col = completion_time,
    into = c("start_time", "end_time"),
    sep = "_"
    )
```

## Splitting Problematic Columns: Specify Separations

```{r 4_separate_funrec_datetime_column_fixed_see_class, eval = FALSE}
# see the data
demo_gathered
```

```{r 4_separate_funrec_datetime_column_fixed_see, echo = FALSE}
# see the data
demo_gathered %>% select(c(1, 6:10))
```

### Converting from Long to Wide Formats with `spread()`

- Useful for paired samples *t*-tests
- Need a key (ID col) and value (scores col)
- If you don't have a value column, create it with `mutate()`!

```{r 4_spread_data_class, eval = FALSE}
demo_gathered %>% 
  mutate(known = 1) %>%
  spread(key = prog_lang, value = known)
```

```{r 4_spread_data, echo = FALSE}
demo_gathered %>% 
  mutate(known = 1) %>%
  spread(key = prog_lang, value = known) %>%
  select(c(1, 7 : 16))
```

# Joins
## Joins

How do we merge two data sets together? We use joins! We have 3 types:

- mutating: creates new columns (e.g. combining demographic data with test data)
- filtering: cuts out observations (e.g. keeping test data for which you have demographic data)
- binding: adding rows or columns (e.g. adding new test scores to existing data)

## Joins: Load Data

```{r 4_load_raw_data, include = FALSE}
lexdec_data <- read_csv("../inputs/lexical_decision_raw_data.csv")
```

```{r 4_load_raw_data_class, eval = FALSE}
lexdec_data <- read_csv("inputs/lexical_decision_raw_data.csv")
```

Subset to only 1 observation for ease of handling (in real life, you wouldn't do this). But we'll cover subsetting in lesson 5.

```{r}
# keep only trials (rows) where the word is ant
lexdec_subset <- lexdec_data %>% filter(word == "ant")
```

```{r 4_peek_raw_data}
lexdec_subset
```

## Joins: Prepare Data

- can join with mismatched names, but best to keep the same column headings for the same variables across data sets. 
- if not, we can use the `rename()` function to change names. 
- in `rename()` assign the new name your old column name.

## Joins: Prepare Data

```{r 4_rename_demo_gathered}
demo_gathered <- rename(demo_gathered, 
                        subject = ID, 
                        native_language = LANGUAGE
                        )
demo_gathered
```

# Mutating Joins
## Full Join

- keeps all data from both data sets. What happens if we try it with our data?

```{r 4_full_join_data_long}
full_join(lexdec_subset, 
          demo_gathered, 
          by = c("subject", "native_language")
          )
```

- multiple rows due to multiple observations

## Full Join

We can fix this by merging by wide data, or by filtering redundant rows after merging.  
- we'll just use wide data as it's easier.

```{r 4_make_tidy_wide_demo_data}
tidy_demo_wide <- demo_gathered %>% 
  mutate(known = 1) %>% # create a value column
  spread(key = prog_lang, value = known) # data to wide format
```

## Full Join

```{r 4_make_tidy_wide_demo_data_output}
# see the output
tidy_demo_wide
```


## Full Join

```{r 4_full_join_class, eval = FALSE}
full_join(lexdec_subset, 
          tidy_demo_wide, 
          by = c("subject", "native_language")
          )
```

```{r 4_full_join, echo = FALSE}
full_join(lexdec_subset, tidy_demo_wide, by = c("subject", "native_language")) %>% 
  select(subject, trial, native_language, progress, correct, RT) %>%
  filter(subject %in% c("23", "24", "A1", "A2", "28"))
```

*Note*: I kept only these 5 subjects to show the important cases.

## Full Join

- 1 row for each participant
- In cases where we don't have data on a subject, we get NAs 

Look at: 

- subject 23, missing native language and trial info: in both data sets, but with missing values
- subject A2, missing trial information. Missing from the lexdec_subset data set, but not the tidy_demo_wide data set

If you're missing from both, you can't be merged!

## Inner Join

- keeps data only present in **both** data sets

- lost participants A1 and A2; A1 wasn't present in the tidy_demo_wide data set, A2 wasn't present in the lexdec_subset data set

```{r 4_inner_join_class, eval = FALSE}
inner_join(lexdec_subset, 
           tidy_demo_wide, 
           by = c("subject", "native_language")
           )
```

```{r 4_inner_join, echo = FALSE}
inner_join(lexdec_subset, tidy_demo_wide, by = c("subject", "native_language")) %>% 
  select(subject, trial, native_language, progress, correct, RT)
```

## Left Join

- only keep the data that is present in the left data set (lexdec_subset) and adds anything that matches up from the right data set (tidy_demo_wide). 

- Here we have participant A1 because they are in the lexdec_subset, even if they are missing from the tidy_demo_wide data set.

```{r 4_left_join_class, eval = FALSE}
left_join(lexdec_subset, 
          tidy_demo_wide, 
          by = c("subject", "native_language")
          )
```

```{r 4_left_join, echo = FALSE}
left_join(lexdec_subset, tidy_demo_wide, by = c("subject", "native_language")) %>% 
  select(subject, trial, native_language, progress, correct, RT)
```

## Right Join

- like the left join, only it keeps everything present in the right data set and anything matching from the left data set

- do not have data on participant A1 because they are not present in the tidy_demo_wide data set

```{r 4_right_join_class, eval = FALSE}
right_join(lexdec_subset, 
           tidy_demo_wide, 
           by = c("subject", "native_language")
           )
```

```{r 4_right_join, echo = FALSE}
right_join(lexdec_subset, tidy_demo_wide, by = c("subject", "native_language")) %>% 
  select(subject, trial, native_language, progress, correct, RT)
```

# Filtering Joins
## Filtering Joins

We can filter data by using joins. 

These next joins don't merge columns, but instead allow us to just subset our data.

## Semi Join

- keep all rows and columns from the left data set where we have matching values in the right data set
- We do not keep the columns from the right data set
- like an `inner_join()` but does not duplicate rows (wide/long data, doesn't matter!)
- only kept data in the lexdec_subset for subjects that were present in both data sets, i.e. not subjects A1 and A2. 

## Semi Join

```{r 4_semi-join_wide}
semi_join(lexdec_subset, 
          tidy_demo_wide, 
          by = c("subject", "native_language")
          )
```

## Anti Join

- the inverse of a semi-join; get all the values from the left table that do not have a match in the right table
- only get participant A1 from the lexdec_subset data set, as we do not have any demographic information on this subject

## Anti Join

```{r 4_anti-join}
anti_join(lexdec_subset, 
          tidy_demo_wide, 
          by = c("subject", "native_language")
          )
```

# Binding Joins
## Binding Joins

- bind rows from separate data sets with the same number of columns using the `bind_rows()`; useful for experiments ran in two parts (same columns, new rows)
- bind columns from separate data sets with the same number of rows using the `bind_cols()`; useful for adding new data to an existing data set where you have the same number of rows

Not covered here -- try it in the Exercise section.

## Checking for Unique and Duplicate Information

We can use a number of functions to check for unique information across two different data sets.

- `intersect()` gives us all the rows in two tables that match exactly; useful for catching duplicates across data sets. 
- `union()` gives us all of the rows from two tables except any duplicates; useful for filtering out duplicates
- `setdiff()` gives us rows from our first data set that aren't present in the second; useful for finding discrepancies

Again, we can get some experience with this in the Exercises section.

# Exercises