---
title: "10 - Reporting Reproducible Research"
output: html_notebook
---

# Introduction and Setup

For these exercises, we will look at the core concepts from this lesson. As always, load the libraries necessary for this lesson.

```{r 10_exercise_libraries, message=FALSE}
library(tidyverse)
library(effsize)
library(TOSTER)
```

For this exercise, we'll look at doing a full analysis using a data set and report from the excellent [Open Stats Lab](https://sites.trinity.edu/osl).

Schroeder and Epley(2015) explored whether you'd be more likely to get a new job after delivering a speech describing your skills, or if you wrote a speech and had a potential employer read it out.

Prediction: A person's speech (i.e., vocal tone, cadence, and pitch) communicates information about intelligence better than written words.
Participants: 39 professional recruiters in one of two conditions. 
Conditions: Listen to audio recording of speech, or read speech aloud.
Dependent Variable: Intellect aggregated across intelligence, competence, and thoughtfulness.

Also rated overall impression (composite of positive and negative), and how likely they would be to hire them (0 = not at all, 10 = extremely likely).

```{r 10_load_data}
# unique URL to the data
id <- "0Bz-rhZ21ShvOei1MM24xNndnQ00" # google file ID

# read in the data
raw_data <- read_csv(
  sprintf(
  "https://docs.google.com/uc?id=%s&export=download", id)
  )
```

## Question 1

Subset the data to the columns relevant to the research question, and make the names all lowercase for consistency. 

Call this, `data_subset`.

```{r 10_question_one}
data_subset <- raw_data %>% 
  select(CONDITION, 
         Intellect_Rating,
         Impression_Rating,
         Hire_Rating
         ) %>%
  mutate(CONDITION = as.factor(CONDITION)) %>%
  rename_all(tolower)
```

## Question 2

Summarise the data, creating means, standard deviations, and ns for each group in the data set. Call this `data_summary`. Put the summary in a table using the `kable()` command.

```{r 10_question_two}
data_summary <- data_subset %>% 
  group_by(condition) %>% 
  summarise(mean_intellect = mean(intellect_rating),
            sd_intellect = sd(intellect_rating),
            n = n()
            )
knitr::kable(data_summary)
```

## Question 3

Create a plot to display the differences across the groups.

```{r 9_question_three}
ggplot(data = data_subset, 
       mapping = aes(x = condition, y = intellect_rating)
       ) +
  geom_boxplot()
```

## Question 4

Test the assumptions of an independent-samples *t*-test.

```{r 10_question_four}
# bartlett test
bartlett.test(intellect_rating ~ condition, data = data_subset)

# qq-plot
ggplot(data = data_subset, mapping = aes(sample = intellect_rating)) +
  stat_qq()
```

**Answer**: The homogeneity of variances assumption is met, as is the assumption of normality; therefore we can procede with the independent-samples *t*-test.

## Question 5

Assuming the test assumptions are met, fit a model to your data, save this as `t_test_output`. For this case, we will just look at the effect of Intellect. Output the test on completion. (Note: do not use `summary()` here, instead just type the name of the object.)

```{r 10_question_five}
t_test_output <- t.test(intellect_rating ~ condition, 
                        paired = F, 
                        data = data_subset
                        )
t_test_output
```

## Question 6

Calculate effect sizes for the *t*-test. The choice of effect size is up to you.

```{r 10_question_six}
cohen.d(intellect_rating ~ condition, 
        paired = F, 
        hedges.correction = T, 
        data = data_subset
        )
```

## Question 7

Perform a TOST on your data evaluating the equivalence against a lower bound of -0.5 and upper bound of 0.5. What does this tell us? (Hint: Use the `with()` function on the data_summary object to access the columns for that data source.)

```{r 10_question_seven}
with(data_summary, 
     TOSTtwo.raw(m1 = mean_intellect[1],
                 m2 = mean_intellect[2],
                 sd1 = sd_intellect[1],
                 sd2 = sd_intellect[2],
                 n1 = n[1],
                 n2 = n[2],
                 low_eqbound = -0.5,
                 high_eqbound = 0.5, 
                 alpha = 0.05, 
                 var.equal = TRUE
                 )
     )
```

**Answer**: The NHST is significant, indicating a statistical difference between the groups. The TOST is non-significant, indicating that the two groups are non-equivalent. This means that the difference between groups is large enough for us to care about them.

## Question 8

A. Fit a linear model to the data. Save this as `lm_test_output`. How does this compare with the *t*-test?
B. Output 95% confidence intervals for the linear model. Why do they differ to that of the *t*-test?

```{r}
lm_test_output <- lm(intellect_rating ~ condition, data = data_subset)
summary(lm_test_output)
confint(lm_test_output)
```



