---
title: "10 - Reporting Reproducible Research"
output: html_notebook
---

# Introduction and Setup

For these exercises, we will look at the core concepts from this lesson. As always, load the libraries necessary for this lesson.

```{r 10_exercise_libraries, message=FALSE}
library(tidyverse)
library(effsize)
library(TOSTER)
```

For this exercise, we'll look at doing a full analysis using a data set and report from the excellent [Open Stats Lab](https://sites.trinity.edu/osl).

Schroeder and Epley(2015) explored whether you'd be more likely to get a new job after delivering a speech describing your skills, or if you wrote a speech and had a potential employer read it out.

Prediction: A person's speech (i.e., vocal tone, cadence, and pitch) communicates information about intelligence better than written words.
Participants: 39 professional recruiters in one of two conditions. 
Conditions: Listen to audio recording of speech, or read speech aloud.
Dependent Variable: Intellect aggregated across intelligence, competence, and thoughtfulness.

Also rated overall impression (composite of positive and negative), and how likely they would be to hire them (0 = not at all, 10 = extremely likely).

```{r 10_load_data}
# unique URL to the data
id <- "0Bz-rhZ21ShvOei1MM24xNndnQ00" # google file ID

# read in the data
raw_data <- read_csv(
  sprintf(
  "https://docs.google.com/uc?id=%s&export=download", id)
  )
```

## Question 1

Subset the data to the columns relevant to the research question, and make the names all lowercase for consistency. 

Call this, `data_subset`.

```{r 10_question_one}

```

## Question 2

Summarise the data, creating means, standard deviations, and ns for each group in the data set. Call this `data_summary`. Put the summary in a table using the `kable()` command.

```{r 10_question_two}

```

## Question 3

Create a plot to display the differences across the groups.

```{r 9_question_three}

```

## Question 4

Test the assumptions of an independent-samples *t*-test.

```{r 10_question_four}

```

**Answer**: 

## Question 5

Assuming the test assumptions are met, fit a model to your data, save this as `t_test_output`. For this case, we will just look at the effect of Intellect. Output the test on completion. (Note: do not use `summary()` here, instead just type the name of the object.)

```{r 10_question_five}

```

## Question 6

Calculate effect sizes for the *t*-test. The choice of effect size is up to you.

```{r 10_question_six}

```

## Question 7

Perform a TOST on your data evaluating the equivalence against a lower bound of -0.5 and upper bound of 0.5. What does this tell us? (Hint: Use the `with()` function on the data_summary object to access the columns for that data source.)

```{r 10_question_seven}

```

**Answer**: 

## Question 8

A. Fit a linear model to the data. Save this as `lm_test_output`. How does this compare with the *t*-test?
B. Output 95% confidence intervals for the linear model. Why do they differ to that of the *t*-test?

```{r}

```



