---
title: "02 - Data Visualisation 2"
subtitle: Basic plotting in ggplot2
author: Glenn Williams
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: ioslides_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(fig.width=5, fig.height=3) 
```

# overview

## Lesson Structure

Thanks to last week's snow, we'll recap on some of the topics from last week.

We'll look at:

- Reading data into R from external sources
- Understanding the ggplot syntax
- Plotting discrete and continuous data
- Changing basic visual features of plots

## Getting Started

First, we'll load the packages necessary for this class. 

```{r 3_load_packages, include = FALSE}
library(tidyverse)
```

```{r 3_load_packages_class, eval = FALSE}
library(tidyverse)
```

## Loading Data

For this recap and the main part of the lessons, we'll use some simulated data of reaction times.

Load the .csv file from the inputs folder. `read_csv()` is a quick way to load data as a tibble.

```{r 3_load_simulated_data, echo = FALSE, message = FALSE}
rt_data <- read_csv("../inputs/rt_data.csv")
```

```{r 3_class_load_simulated_data, eval = FALSE}
rt_data <- read_csv("inputs/rt_data.csv")
```

## Understanding the Data

Let's see what the data looks like:

```{r 3_data_peek}
rt_data
```

# ggplot

## Basics of Plotting

Every `ggplot` call must have a data argument!

```{r 3_boxplot}
ggplot(data = rt_data)
```

... but that's not interesting. We've just made a canvas for our data.

## Mapping your Data onto the Plot

```{r 3_no_geom}
ggplot(data = rt_data) +
  geom_boxplot(mapping = aes(x = response_condition, 
                             y = reaction_time)
               )
```

This is the most basic boxplot you can make in ggplot.

## Customising our Plots

### Adding Colour across Variables

```{r 3_boxplot_colour}
ggplot(data = rt_data) +
  geom_boxplot(mapping = aes(x = response_condition, 
                             y = reaction_time),
               colour = "red"
               )
```

## Customising our Plots

### Adding Colour by Variables

Notice, we get a legend when we add colours by variables.

```{r 3_boxplot_colour_by_var}
ggplot(data = rt_data) +
  geom_boxplot(mapping = aes(x = response_condition, 
                             y = reaction_time, 
                             colour = response_condition))
```

## Customising our Plots

What if we want to expand our y-axis? Set your limits!

```{r 3_boxplot_colour_by_var_ylim}
ggplot(data = rt_data) +
  geom_boxplot(mapping = aes(x = response_condition, 
                             y = reaction_time, 
                             colour = response_condition)
               ) +
  coord_cartesian(ylim = c(0, 600))
```

# Geoms

## Exploring Other Geoms

- `geom_point()`: needs x and y in in `aes()`
- `geom_bar()`: needs x and y in `aes()`
- `geom_violin()`: needs x and y in `aes()`
- `geom_density()` : only needs x values in `aes()`
- `geom_histogram()` : only needs x values in `aes()`

## Density Plots

```{r 3_density}
ggplot(data = rt_data, aes(x = reaction_time)) +
  geom_density(mapping = aes(fill = response_condition),
               alpha = 0.6
               )
```

## Bar Plots

Good for showing counts: Here by gender and response condition.

```{r 3_barplot}
ggplot(data = rt_data, aes(x = gender)) +
  geom_bar(mapping = aes(fill = response_condition))
```

... not so great for non-count data.

## Pirate Plots! Rrrrrrrrrrr

Great for the the indecisive. Shows:

- Bars of mean scores
- Box of 95% confidence interval around the mean
- Violins for density and distribution of scores
- Raw data points

Let's load it!

```{r 3_load_ggpirate}
# install.packages("devtools")
# devtools::install_github("mikabr/ggpirate")
library(ggpirate)
```

## Pirate Plots! Rrrrrrrrrrr

```{r 3_pirate_plot}
ggplot(data = rt_data, 
       mapping = aes(x = response_condition, y = reaction_time)
       ) +
  geom_pirate()
```

## Saving Plots as Objects

- Avoid retyping/rerunning the same code: Assign to an object!

```{r 3_pirate_plot_object}
pirate_plot <- ggplot(data = rt_data, 
                      mapping = aes(x = response_condition, 
                                    y = reaction_time)
                      )
pirate_plot # enter name in console to see the plot!
```

## Saving Plots as Objects

This looks neater now...

```{r 3_pirate_plot_object_with_geom}
pirate_plot + geom_pirate()
```

## Extreme Customisation!

```{r 3_pirate_plot_object_custom_code, eval = FALSE}
pirate_plot + 
  geom_pirate(mapping = aes(fill = response_condition, 
                            colour = response_condition)
              ) +
  labs(x = "Response Condition", y = "Reaction Time (ms)") +
  scale_x_discrete(labels = c("Match", "Mismatch")) +
  theme_classic() +
  scale_fill_manual(values = c("red", "blue")) +
  scale_colour_manual(values = c("red", "blue")) +
  geom_hline(yintercept = 480, linetype = 2) + 
  annotate(geom = "text", x = 1, y = 500, label = "Very Slow Performance")
```

## Extreme Customisation!

```{r 3_pirate_plot_object_custom_graph, echo = FALSE}
pirate_plot + 
  geom_pirate(mapping = aes(fill = response_condition, 
                            colour = response_condition)
              ) +
  labs(x = "Response Condition", y = "Reaction Time (ms)") +
  scale_x_discrete(labels = c("Match", "Mismatch")) +
  theme_classic() +
  scale_fill_manual(values = c("red", "blue")) +
  scale_colour_manual(values = c("red", "blue")) +
  geom_hline(yintercept = 480, linetype = 2) + 
  annotate(geom = "text", x = 1, y = 500, 
           label = "Very Slow Performance"
           )
```

Why no legend? ggpirate defaults to remove that (not necessary for discrete data).

# Faceting

## Faceting

- A useful way to make multiple plots on one canvas
- Useful for splitting categories in your data

For splitting by one category:

```{r facet_wrap_form, eval = FALSE}
facet_wrap(~ category)
```

For multiple categories:

```{r facet_grid_form, eval = FALSE}
facet_grid(category_one ~ category_two)
```

Both cases can take two categories, but the grid layout is more interpretable for multiple categories.

## Facet Examples

```{r facet_grid_example}
ggplot(data = rt_data, 
       mapping = aes(x = reaction_time)
       ) +
  geom_density() +
  facet_wrap(~ response_condition)
```

## Facet Examples

```{r facet_wrap_example}
ggplot(data = rt_data, 
       mapping = aes(x = reaction_time)
       ) +
  geom_density() +
  facet_grid(gender ~ response_condition)
```

# Calculating Statistics

## Means and SE

```{r means_se_plot}
ggplot(data = rt_data, 
       mapping = aes(
         x = response_condition,
         y = reaction_time, 
         fill = response_condition)
       ) + 
  stat_summary(fun.y = mean, geom = "bar") +
  stat_summary(fun.data = "mean_se", geom = "errorbar", width = 0.25)
```

## Model Fits

Defaults to loess smooth.

```{r loess_fit_code, eval = FALSE}
# filter extreme values
starwars <- filter(starwars, mass < 160)

# plot
ggplot(data = starwars, 
       mapping = aes(x = mass, y = height)
       ) + 
  geom_point() + 
  geom_smooth()
```

## Model Fits

```{r loess_fit_plot, echo = FALSE}
# filter extreme values
starwars <- filter(starwars, mass < 160)

# plot
ggplot(data = starwars, 
       mapping = aes(x = mass, y = height)
       ) + 
  geom_point() + 
  geom_smooth()
```

## Model Fits

Fitting a linear model

```{r lm_fit}
ggplot(data = starwars, 
       mapping = aes(x = mass, y = height)
       ) + 
  geom_point() + 
  geom_smooth(method = "lm", formula = "y ~ x")
```

# Combining Plots

## Combining Plots

```{r lm_fit_saved}
linear_plot <- 
  ggplot(data = starwars, 
       mapping = aes(x = mass, y = height)
       ) + 
  geom_point() + 
  geom_smooth(method = "lm", formula = y ~ x)
```

```{r quad_fit_saved}
quadratic_plot <-
  ggplot(data = starwars, 
       mapping = aes(x = mass, y = height)
       ) + 
  geom_point() + 
  geom_smooth(method = "lm", formula = y ~ poly(x, 2))
```

*Note*: We can save on retyping everything above geom smooth by saving the arguments to a basic `plot`, then adding that + `geom_smooth()` to our quadratic/linear plot names.

## Combining Plots

```{r combining_plots, messages = FALSE}
library(cowplot)

combined_plots <- plot_grid(linear_plot, 
                            quadratic_plot, 
                            labels = c("A", "B"), # label plots
                            align = "h" # align axes
                            )
```

## Combining Plots

```{r return_combined_plots}
combined_plots
```

# Saving

## Saving

Define the location and name of the file in the `filename` argument.

Define the object to save in your plot argument.

*Note*: Here, we use **relative** file paths. We tell R, in the current folder, find the outputs folder, within that (/) save the image as a .png by the name starwars_mass_by_height.

```{r ggsave, eval = FALSE}
ggsave(filename = "outputs/starwars_mass_by_height.png", 
       plot = combined_plots
       )
```

# Exercises