---
title: "04 - Data Manipulation One"
output: html_notebook
---

# Introduction and Setup

For these exercises, we will look at the core concepts from this lesson. We'll also get some hands-on experience with binding joins and checking for duplicates, two concepts that we've touched on but not went into much detail.

For these exercises we'll use some toy data sets; ex_demo_data, which has demographic information on 6 participants, and ex_test_data, which has IQ test scores for 6 participants. Crucially, the first data set has some missing values, and the second has the same participant tested twice.

```{r 4_exercise_libraries_and_data, message=FALSE}
# load the tidyverse
library(tidyverse)

# demographic data
ex_demo_data <- tibble(
  subject = seq(1: 6),
  height = c(NA, 170, 160, 165, NA, 180),
  weight = c(70, 65, 80, NA, 77, 90),
  age = c(18, 19, 19, NA, 22, 28)
  )

# IQ test scores
ex_test_data <- tibble(
  subject = c(1, 3, 4, 4, 5, 6, 7),
  IQ = c(150, 160, 155, 155, 190, 120, 140)
  )
```

First, you should check out both data sets:

```{r 4_check_data}
ex_demo_data
ex_test_data
```


# Long and Wide Data

## Question 1

Put the `ex_demo_data` into a long format with three columns: subject, measurement_id, and measurement. The measurement column should contain the scores for the height, weight, and age of the participants. The measurement_id column should contain text specifying which measurement belongs to which variable (height, weight, or age). Assign this to the variable `long_data` and return this table of data.

```{r 4_question_one_answer}

```

## Question 2

Turn your newly created `long_data` back into a wide format.

```{r 4_question_two_answer}

```

# Uniting and Separating Columns

Here we have some messy data where we have two values for two variables in one column; `height_weight`.

```{r 4_make_messy_data}
messy_demo_data <- unite(ex_demo_data, 
                         "height_weight", 
                         c("height", "weight"), 
                         sep = "_"
                         )
messy_demo_data
```

## Question 3

Separate the messy columns into two tidy columns for height and weight. Should you convert the values when separating the column? If so, why? Write your answer below.

```{r 4_question_three_answer}

```

**Answer**:

# Mutating Joins

## Question 4

Join the `ex_demo_data` and `ex_test_data` together by subject number, keeping only data with a match in `ex_test_data`.

```{r 4_question_four_answer}

```

## Question 5

Join the `ex_demo_data` and `ex_test_data` together by subject number, keeping only data with a match in `ex_demo_data`.

```{r 4_question_five_answer}

```

## Question 6

Why do we get different results in question 4 and question 5? Write your answer below.

**Answer**:

# Filtering Joins

## Question 7

Return all of the values from `ex_demo_data` that have a match in `ex_test_data`. Look at subject 4, why do we get a different result to that from question 5? Look at the columns returned, why does this differ from question 5? Write your answer below.

```{r 4_question_seven_answer}

```

**Answer**:

# Binding Joins

Here we have some new data looking at the demographic scores for new subjects. We also have another rating for all of the participants from our study and we want to add this to the demographic data.

```{r 4_make_binding_joins_data}
new_demographics <- tibble(
  subject = c(9, 10),
  height = c(170, 190),
  weight = c(76, 85),
  age = c(40, 59)
  )

eye_colour <- tibble(
  eye_colour = sample(c("blue", "brown", "green"), 
                      size = 8, 
                      replace = TRUE
                      )
  )
```

## Question 8

Add the rows from `new_demographics` to `ex_demo_data`. Assign this to `all_demo_data` and return this table.

```{r 4_question_eight_answer}

```

## Question 9

Add the eye colour column to the `all_demo_data` table. Why did we not have a subject identifier in the `eye_colour` data set? Can you predict the result if we did have this information? Write your answer below.

```{r 4_question_nine_answer}

```

**Answer**:

# Checking for Duplicates

We have some new test data below.

```{r 4_make_duplicates_data}
extra_test_data <- tibble(
  subject = c(1, 9, 10),
  IQ = c(150, 156, 179)
  )
extra_test_data
```

## Question 10

Return rows with duplicates from the `ex_test_data` and `extra_test_data` data sets.

```{r 4_question_ten_answer}

```
