---
title: "05 - Data Manipulation Two"
output: html_notebook
---

# Introduction and Setup

For these exercises, we will look at the core concepts from this lesson.

For these exercises we'll use a simulated data set I created in the **data_generation** folder. This data was saved as a tibble in the RData file, so we just use `load()` to open this data into R.

```{r 5_exercise_libraries_data, message=FALSE}
# load libraries
library(tidyverse)
library(lubridate)

# load data
load("inputs/sim_data.RData")
```

This data set looks at reaction times in spotting a target when people have had caffeine or not, and when they respond with their dominant or non-dominant hand. This data set is entirely made up, so we can't be sure how realistic these results are.

First, you should take a look at your data to understand it. A simple first step is to see what the data table itself looks like.

```{r understand_data}
glimpse(data)
```

After this, we should make a more detailed plot to understand the general trends in the data.

## Question 1

Make a plot of your data which looks at the relationship beween **caffeine** and **DV**. Split this plot into two to show the differences across the two response conditions.

```{r 5_question_one_answer}

```
What are the general trends in the data?

## Question 2

Subset your data to remove the ages from the data set.

```{r 5_question_two_answer}

```

## Question 3

Rename the **DV** column to something more informative, like **reaction_time**

```{r 5_question_three_answer}

```

## Question 4

What if we care about differences in the ages? Let's assume we have a prediction that caffeine only has an effect on those above 30 years of age. Subset your data to just those above 30 years of age.

```{r 5_question_four_answer}

```

## Question 5

Rearrange the data set by age, starting from the **highest** age.

```{r 5_question_five_answer}

```

## Question 6

Calculate mean centered scores for each subject and add these to a new column called **DV_c** (DV, centered). The formula for this is `subjects_score - mean(all_scores)`. Can you work out what the mean should (approximately) be?

```{r 5_question_six_answer}

```
**Answer**:

## Question 7

Let's assume we have a prediction that response times should be slower above 30 years of age. Create a new column, **age_group** that puts participants into two groups **30_or_under**, and **above_30**.

*Hint*: Look up the `ifelse()` function using `?ifelse()` to see how you can use logical operations to achieve this.

```{r 5_question_seven_answer}

```

## Question 8

Calculate the mean, standard deviation, and number of observations for each group.

```{r 5_question_eight_answer}

```

## Question 9

Calculate the mean, standard deviation, and number of observations for each group, excluding those with the 3 highest ages from each group.

```{r 5_question_nine_answer}

```

## Question 10 

Take the data and do this all together:

1. Rename the DV column to **response_time**.
2. Remove any observations where the age of the participant is above 60.
3. Combine the two columns, age and caffeine, into one column called condition. Hint: Use `paste()` here. Use an underscore separator for the condition names.
4. Remove the caffeine and response columns and reorder the data so we have subject first, followed by age, condition, and response_time.
5. Calculate mean, standard deviation, and n for condition on the response time column. Call your new column names anything sensible.

```{r 5_question_ten_answer}

```

## Question 11

Was there any point in us combining the two factors into a single condition column? Do the same process above without making a summary of the data. Feed this data into a boxplot, with condition on the x-axis, and response_time on the y-axis.

```{r 5_question_11_answer}

```

