---
title: "07 - Advanced Statistical Tests Answers"
output: html_notebook
---

# Introduction and Setup

For these exercises, we will look at the core concepts from this lesson. Here, we'll use a simulated data set that's saved as a .csv in the inputs folder for this class.

```{r 7_exercise_libraries, message=FALSE}
library(tidyverse)
# install.packages("car") # install if necessary (once per machine)
library(car)
library(ggpirate)
```

Next, we'll load the data set for this exercise.

```{r message = FALSE}
factorial_data <- read_csv("inputs/factorial_data.csv")
```

## Question 1

Aggregate the data by subject such that you have 1 score per subject for the mean of the column Y across both A and B columns. Call the Y column mean_Y. Assign this to the object **data_agg**.

```{r 7_question_one}
data_agg <- factorial_data %>% 
  group_by(subj_id, A, B) %>% 
  summarise(mean_Y = mean(Y))
```

## Question 2

Generate summary statistics for the data set and a pirate plot of the data set. We want means, SDs, and ns for the table of summary statistics.

```{r 7_question_two_a}
data_agg %>% 
  group_by(A, B) %>% 
  summarise(mean = mean(mean_Y),
            sd = sd(mean_Y),
            n = n()
            )
```

```{r 7_question_two_b}
data_agg %>% 
  ggplot(mapping = aes(x = A, y = mean_Y)) +
  facet_wrap(~ B) +
  geom_pirate()
```

## Question 3

Define your contrasts such that you have main effects and interactions from any model fits. You may need to check your data types for this to work properly. 
*Hint*: use `mutate()` after the `ungroup()` function

```{r 7_question_three}
data_agg <- data_agg %>%
  ungroup() %>%
  as.tibble() %>%
  mutate(A = factor(A),
         B = factor(B)
         )

contrasts(data_agg$A) <- contr.sum
contrasts(data_agg$B) <- contr.sum
```

## Question 4

Fit the data set checking for all main effects and interactions.

```{r 7_question_four}
aov_fit <- aov(mean_Y ~ A * B, data = data_agg)
summary(aov_fit)
```

## Question 5

Refit your model from Question 5 using type-III sum of squares. Did that make any difference to your sum of squares? Why or why not?

```{r 7_question_five}
Anova(aov_fit, type = "III")
```

**Answer**: The sum of squares were unaffected as we had a balanced data set.

## Question 6

Explore the parameter estimates from your ANOVA model, and briefly report your findings in text (1 sentence max).

```{r 7_question_six}
summary.lm(aov_fit)
```

**Answer**: We found a significant main effect of B on (*t* = 2.93, *p* = .004), however the main effect of A, and the interaction between A and B was non-significant (both *p*s > .05).
