---
title: "07 - Advanced Statistical Tests Answers"
output: html_notebook
---

# Introduction and Setup

For these exercises, we will look at the core concepts from this lesson. Here, we'll use a simulated data set that's saved as a .csv in the inputs folder for this class.

```{r 7_exercise_libraries, message=FALSE}
library(tidyverse)
# install.packages("car") # install if necessary (once per machine)
library(car)
library(ggpirate)
```

Next, we'll create the data set for this exercise.

```{r 7_exercise_data, message = FALSE}
factorial_data <- read_csv("inputs/factorial_data.csv")
```

## Question 1

Aggregate the data by subject such that you have 1 score per subject for the mean of the column Y across both A and B columns. Call the Y column mean_Y. Assign this to the object **data_agg**.

```{r 7_question_one}

```

## Question 2

Generate summary statistics for the data set and a pirate plot of the data set. We want means, SDs, and ns for the table of summary statistics.

```{r 7_question_two_a}

```

```{r 7_question_two_b}

```

## Question 3

Define your contrasts such that you have main effects and interactions from any model fits. You may need to check your data types for this to work properly. 
*Hint*: use `mutate()` after the `ungroup()` function

```{r 7_question_three}

```

## Question 4

Fit the data set checking for all main effects and interactions.

```{r 7_question_four}

```

## Question 5

Refit your model from Question 5 using type-III sum of squares. Did that make any difference to your sum of squares? Why or why not?

```{r 7_question_five}

```

**Answer**: 

## Question 6

Explore the parameter estimates from your ANOVA model, and briefly report your findings in text (1 sentence max).

```{r 7_question_six}

```

**Answer**: 
