---
title: "Lesson 2 - Data Visualisation"
output: html_notebook
---

# Load libraries and Data

Load the libraries and data necessary to complete the exercises associated with this lesson. You may have to install the `languageR` library if you don't have it installed already.

```{r load_libraries_and_data, messages = FALSE}
library(tidyverse)
library(ggpirate)
library(cowplot)

# install.packages("languageR")
library(languageR)

data <- lexdec %>% select(c("Subject", "Sex", "NativeLanguage",
                            "Trial", "Word", "Class", "RT")
                          )
```

# Exercises

Enter your code in the code blocks in response to the [exercises for Chapter 2 of r4psych](http://glennwilliams.me/r4psych/data-visualisation.html#exercises). Run your code and then save your document once completed.

## Question 1

```{r Question 1}

```

## Question 2

```{r Question 2}

```

## Question 3

```{r Question 3}

```

## Question 4

```{r Question 4}

```

## Question 5

```{r Question 5}

```

## Question 6

```{r Question 6}

```

## Question 7

```{r Question 7}

```

## Question 8

```{r Question 8}

```

## Question 9

```{r Question 9}

```

## Question 10

```{r Question 10}

```
